Topological Equivariant Artist Model for Visualization Library Architecture

### 2024-12-04 paper/main.tex
related work

\begin{figure}[]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{This weather station data has multiple embedded continuities - points at each time and position, timeseries at each position, and maps at each time. The corresponding visualizations - bar chart, timeseries, and map - each preserve the continuity of the subset of the data they visualize such that there is continuous map from the visualization (continuity) to the data (continuity).}
%
  \label{fig:related-work:continuity:ktypes}
\end{figure}

For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner because it introduces ambiguity over which part of the line maps back to the data. A map that by definition has continous maps between two spaces, such as data and graphics, is called a \textit{homeomorphism}\cite{riehlCategoryTheoryContext}:

\begin{definition}
  A function $f$ is a $homeomorphism$ if it is bijective, continous, and has a continous inverse function $f^{-1}$.
\end{definition}

Therefore the the bar plot, line plot, and heatmap in \autoref{fig:related-work:continuity:ktypes} have a homemeomorphic relationship to the 0D points, 1D timeseries, and 2D map continuities embedded in the continuous 3 dimensional surface encapsulating time and position. Using homeomorphism to test whether continuity is preserved generalizes Bertin's codification of how the topology of observations must match the class of representation (i.e. point , line, area) \cite{bertinSemiologyGraphicsDiagrams2011} and Wilkinson's assertion that connectivity be preserved \cite{wilkinsonGrammarGraphics2005}.

@@

\begin{definition}
A \textbf{topology} $\mathcal{T}$ on a space $X$ is a collection of open sets\footnote{Open sets (open subsets) are a generalization of open intervals to n dimensional spaces. For example, an open ball is the set of points inside the ball and excludes points on the surface of the ball. \cite{weissteinOpenSet,bradleyTopologyVsTopology}} \openset\ of $X$. This collection has the properties that the empty set $\varnothing$ and $\dbase$ are in $\mathcal{T}$, the union of open sets in $\mathcal{T}$ is also in $\mathcal{T}$, and any finite intersection of open sets in $\mathcal{T}$ is also in $\mathscr{T}$. \cite{bradleyTopologyCategoricalApproach2020}
\end{definition}

@@

Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict their API to choices that that are appropriate for the domain. For example, a tabular topological structure of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanities ImagePlot\cite{studiesCulturevisImageplot2021} plugin assume that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph-like structure. By assuming the structure of their data, these domain specific libraries can provide more cohesive interfaces for a much more limited set of visualization algorithms than the building block libraries offer.

>>>>

\begin{figure}[]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{This weather station data has multiple embedded continuities - points at each time and position, timeseries at each position, and maps at each time. The corresponding visualizations - bar chart, timeseries, and map - each preserve the continuity of the subset of the data they visualize by not introducing or leaving out values and preserving the relative positioning of continuous values.}
%
  \label{fig:related-work:continuity:ktypes}
\end{figure}

For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner because it introduces ambiguity over which part of the line maps back to the data. A map that by definition has continuous maps between the input and output spaces, such as data and graphics, is called a \textit{homeomorphism}\cite{riehlCategoryTheoryContext}:

\begin{definition}
  A function $f$ is a $homeomorphism$ if it is bijective, continuous, and has a continuous inverse function $f^{-1}$.
\end{definition}

The bar plot, line plot, and heatmap in \autoref{fig:related-work:continuity:ktypes} have a homemeomorphic relationship to the 0D points, 1D timeseries, and 2D surface continuities embedded in the continuous 3 dimensional surface encapsulating time and position because each point of the visualization maps back into a point in its corresponding indexing space in the cube. Using homeomorphism to test whether continuity is preserved formalizes Bertin's codification of how the topology of observations matches the class of representation (i.e. point , line, area) \cite{bertinSemiologyGraphicsDiagrams2011} and Wilkinson's assertion that connectivity must be preserved \cite{wilkinsonGrammarGraphics2005}.

@@

\begin{definition}
A \textcolor{base}{\textbf{topology}} $\mathcal{T}$ on a space $X$ is a collection of open sets\footnote{Open sets (open subsets) are a generalization of open intervals to n dimensional spaces. For example, an open ball is the set of points inside the ball and excludes points on the surface of the ball. \cite{weissteinOpenSet,bradleyTopologyVsTopology}} \openset\ of $X$. This collection has the properties that the empty set $\varnothing$ and $\dbase$ are in $\mathcal{T}$, the union of open sets in $\mathcal{T}$ is also in $\mathcal{T}$, and any finite intersection of open sets in $\mathcal{T}$ is also in $\mathscr{T}$. \cite{bradleyTopologyCategoricalApproach2020}
\end{definition}

@@

Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict their API to choices that are appropriate for the domain. For example, a tabular topological structure of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanities ImagePlot\cite{studiesCulturevisImageplot2021} plugin assume that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph-like structure. By assuming the structure of their data, these domain specific libraries can provide more cohesive interfaces for a much more limited set of visualization algorithms than the building block libraries offer.
